Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-11 main.tex
decomp

@@

\newcommand\aiD {\mathsf{ai}}
\newcommand\faD {\forall}
\newcommand\exD {\exists}
\newcommand\tttD {\ttt}
\renewcommand\wD {\mathsf{w}}
\renewcommand\cD {\mathsf{c}}
\renewcommand\acD {\mathsf{ac}}
\newcommand\mfaD {\mathsf{m_\forall}}
\newcommand\mexD {\mathsf{m_\exists}}

@@

\begin{thm}
  A formula $A$ is a theorem of first-order logic iff the sequent $\sqn A$ is derivable in $\FOLK$.
\end{thm}

@@

For a formula $A$, we can define its
associated fograph $\graphof{A}$ inductively by:
\begin{equation*}
  \begin{array}{cc}
    \graphof{\alpha} = \single \alpha & 
    \graphof{\forall x . A} = \single x + \graphof{A} \\[.5ex]
    \graphof{A \cor B} = \graphof{A} + \graphof{B} &
    \graphof{\exists x . A} = \single x \times \graphof{A} \\[.5ex]
    \graphof{A \cand B} = \graphof{A} \times \graphof{B} &

@@

\begin{proof}
  \todo{}
  By a straightforward induction on $A$.
\end{proof}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}
  \begin{center}
    \framebox{%
      $
      \begin{array}{@{}c@{}}
        \dbox{%
          $
          \begin{array}{c}
            \vlinf{\faD}{}{\Scons{\forall x.\ttt}}{\Scons\ttt}
            \qquad
            \vlinf{\aiD}{}{\Scons{a\vlor\cneg a}}{\Scons\ttt}
            \qquad
            \vlinf{\tttD}{}{\Scons{A\vlan\ttt}}{\Scons A}
            \\ \\[-1ex]
            \vlinf{\sw}{}{\Scons{(A\vlan B)\vlor C}}{\Scons{A\vlan (B\vlor C)}}
            \qquad
            \vlinf{\mix}{}{\Scons{A\vlor B}}{\Scons{A\vlan B}}            
            \\ \\[-1ex]
            \vlinf{\exD}{}{\Scons{\exists x.A}}{\Scons{A\ssubst x t}}
            \qquad
            \vlinf{\fequ}{\text{(where $A\fequ B$)}}{\Scons{A}}{\Scons{B}}
          \end{array}
          $
        }
        \\ \\[-1ex]
        \vlinf{\mfaD}{}{\Scons{\forall x.(A\vlor B)}}{\Scons{(\forall x.A)\vlor(\forall x.B)}}
        \qquad
        \vlinf{\mexD}{}{\Scons{\exists x.(A\vlor B)}}{\Scons{(\exists x.A)\vlor(\exists x.B)}}        
        \\ \\[-1ex]
        \vlinf{\me}{}{\Scons{(A\vlor B)\vlan(C\vlor D)}}{\Scons{(A\vlan C)\vlor(B\vlan D)}}
        \\ \\[-1ex]
        \vlinf{\wD}{}{\Scons{A\vlor B}}{\Scons{A}}
        \qquad
        \vlinf{\acD}{}{\Scons{a}}{\Scons{a\vlor a}}
      \end{array}
      $
    }
  \end{center}
  \caption{Deep inference systems $\FOKS$ (all rules) and $\FOMLS$ (rules in the dashed box)}
  \label{fig:KS1}
\end{figure}

@@

We consider here only the cut-free fragment, as cut-elimination for
deep inference systems has already been discussed
elsewhere~\cite{brunnler:06:herbrand,alertubella:guglielmi:18}.
\juihsuan{references added. Please check if they are the right ones.}
\footnote{In the deep
inference literature, the cut-free fragment is also called the
\emph{down-fragment}. But as we do not discuss the \emph{up-fragment}
here, we omit the down-arrows $\downarrow$ in the rule names.}

@@

@@

We are now ready to see the main results of this paper.

>>>>

\newcommand{\sysS}{\mathsf{S}}
\newcommand{\Deri}{\Phi}

@@

\newcommand\aiD {\mathsf{ai}}
\newcommand\faD {\forall}
\newcommand\exD {\exists}
\newcommand\tttD {\ttt}
\renewcommand\wD {\mathsf{w}}
\newcommand\wlD {\mathsf{w_l}}
\newcommand\wrD {\mathsf{w_r}}
\renewcommand\cD {\mathsf{c}}
\renewcommand\acD {\mathsf{ac}}
\newcommand\mfaD {\mathsf{m_\forall}}
\newcommand\mexD {\mathsf{m_\exists}}

@@

\begin{thm}
  $\FOLK$ is sound and complete for first-order logic.
\end{thm}

@@

For a formula $A$, we can define its
associated fograph $\graphof(A)$ inductively by: 
\begin{equation*}
  \begin{array}{cc}
    \graphof{\alpha} = \single \alpha & 
    \graphof{\forall x . A} = \single x + \graphof{A} \\[.5ex]
    \graphof{A \cor B} = \graphof{A} + \graphof{B} &
    \graphof{\exists x . A} = \single x \times \graphof{A} \\[.5ex]
    \graphof{A \cand B} = \graphof{A} \times \graphof{B} &

@@

\begin{proof}
  \todo{}
\end{proof}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}
  \begin{center}
    \framebox{%
      $
      \begin{array}{@{}c@{}}
        \dbox{%
          $
          \begin{array}{c}
            \vlinf{\faD}{}{\Scons{\forall x.\ttt}}{\Scons\ttt}
            \qquad
            \vlinf{\aiD}{}{\Scons{a\vlor\cneg a}}{\Scons\ttt}
            \qquad
            \vlinf{\tttD}{}{\Scons{A\vlan\ttt}}{\Scons A}
            \\ \\[-1ex]
            \vlinf{\sw}{}{\Scons{(A\vlan B)\vlor C}}{\Scons{A\vlan (B\vlor C)}}
            \qquad
            \vlinf{\mix}{}{\Scons{A\vlor B}}{\Scons{A\vlan B}}            
            \\ \\[-1ex]
            \vlinf{\exD}{}{\Scons{\exists x.A}}{\Scons{A\ssubst x t}}
            \qquad
            \vlinf{\fequ}{\text{(where $A\fequ B$)}}{\Scons{A}}{\Scons{B}}
          \end{array}
          $
        }
        \\ \\[-1ex]
        \vlinf{\mfaD}{}{\Scons{\forall x.(A\vlor B)}}{\Scons{(\forall x.A)\vlor(\forall x.B)}}
        \qquad
        \vlinf{\mexD}{}{\Scons{\exists x.(A\vlor B)}}{\Scons{(\exists x.A)\vlor(\exists x.B)}}        
        \\ \\[-1ex]
        \vlinf{\me}{}{\Scons{(A\vlor B)\vlan(C\vlor D)}}{\Scons{(A\vlan C)\vlor(B\vlan D)}}
        \\ \\[-1ex]
        \vlinf{\wrD}{}{\Scons{A\vlor B}}{\Scons{A}}
        \qquad
        \vlinf{\wlD}{}{\Scons{B\vlor A}}{\Scons{A}}
        \qquad
        \vlinf{\acD}{}{\Scons{a}}{\Scons{a\vlor a}}
      \end{array}
      $
    }
  \end{center}
  \caption{Deep inference systems $\FOKS$ (all rules) and $\FOMLS$ (rules in the dashed box)}
  \label{fig:KS1}
\end{figure}

@@

We consider here only the cut-free fragment, as cut-elimination for
deep inference systems has already been discussed
elsewhere~\cite{brunnler:,alertubella:}.\footnote{In the deep
inference literature, the cut-free fragment is also called the
\emph{down-fragment}. But as we do not discuss the \emph{up-fragment}
here, we omit the down-arrows $\downarrow$ in the rule names.}

@@

We write $\vlder{\sysS}{\Deri}{A}{B}$ to denote a derivation $\Deri$
from $B$ to $A$ using the rules from system $\sysS$. A formula $A$ is
\bfit{provable} in a system $\sysS$ if there is a derivation in
$\sysS$ from $\ttt$ to $A$.

In the course of this paper we are also going to make use of the
general (non-atomic) version of the contraction rule:
\begin{equation*}
  \vlinf{\cD}{}{\Scons{A}}{\Scons{A\vlor A}}  
\end{equation*}

@@

We are now ready to see the main results of this paper. We only state
them here and give the proofs in the later sections of the paper. The
first one is routine and expected, but needs to be proved nonetheless:

\begin{thm}\label{thm:KS1}
  $\FOKS$ is sound and complete for first-order logic.
\end{thm}

Our second results is more surprising, as it is a very strong
decomposition result for first order logic.

\begin{thm}\label{thm:decomposition}
  For every derivation $\vlder{\FOKS}{\Deri}{A}{\ttt}$ there are formulas $A_1,\ldots,A_5$, such that there is a derivation:
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wlD,\wrD}}{}{A}{
        \vlde{\set{\acD}}{}{A_1}{
          \vlde{\set{\me,\mfaD,\mexD,\fequ}}{}{A_2}{
            \vlde{\set{\exD}}{}{A_3}{
              \vlde{\set{\sw,\mix,\fequ}}{}{A_4}{
                \vlde{\set{\faD,\aiD,\tttD}}{}{A_5}{
                  \vlhy{\ttt}}}}}}}}
  \end{equation*}
\end{thm}

This theorem is stronger than the existing decompositions for
first-order logic, which either separated only atomic contraction and
atomic weakening~\cite{brunnler:phd} or only
contraction~\cite{ralph:phd} or only the quantifiers in form of a
Herbrand theorem~\cite{brunnler:06:locality,ralph:phd}.

There is a weaker version of Theorem~\ref{thm:decomposition} that will
also be useful:

\begin{thm}\label{thm:decompositionA}
  For every derivation $\vlder{\FOKS}{\Deri}{A}{\ttt}$ there is a formula $A_1$, such that there is a derivation:
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wlD,\wrD,\cD}}{}{A}{
        \vlde{\FOMLS}{}{A_1}{
          \vlhy{\ttt}}}}}}}}
  \end{equation*}
\end{thm}
